Modal Verification Patterns for Systems

### 2025-06-02 paper.tex
ready for dali

%\bibliography{references/paper}
\begin{document}
%
\title{Safe Deferred Memory Reclamation with Types}
%
%\titlerunning{RCU Types}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Ismail Kuru \orcidID{0000-0002-5796-2150} \and
Colin S. Gordon \orcidID{0000-0002-9012-4490}}
%%%Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{I. Kuru et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Drexel University \\ %\and
%%%Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{\{ik335,csgordon\}@drexel.edu}}%\\
%%%\url{http://www.springer.com/gp/computer-science/lncs} \and
%%%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%%%\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
\section{Introduction}
\label{sec:intro}
Systems verification has attracted much attention from both the verification community and the systems community over the past couple of decades. %There are two main streams of verification efforts on systems. In one class of verification, \emph{automated} verification tools are used. They allow people to specify systems via specific annotation languages they offer, and incorporate translation of these specifications into SMT-like formulas for which proofs are searched by the SMT solvers at the backend ~\cite{}. 
Most of this effort has been build on top of \emph{program logics}~\cite{reynolds2002separation,hoare1969axiomatic,smans2012implicit,pratt1976semantical},
which are formal logics of reasoning about program behaviour --- given any starting state
satisfying a particular precondition, running a certain code segment of interest should not fail and should produce a final state satisfying a desired postcondition.
This requires an interacting family of pieces, including a specification language
to describe those desired pre- and postconditions, as well as rules of the logic to deal with
how program operations like variable or register assignments and memory accesses interact with pre- and postconditions.
Since most low-level systems are stateful, \emph{separation logic} has been the focus 
of most recent work.
%has been either the foundational or direct choice of logic in these verification efforts. 
%\Colin{Talking about separation logic makes lots of sense. Do we care about Iris/Coq/Lean/etc. vs. automation? Using the modality idea
%for something like Dafny would certainly look different from the Iris work, but can't it work?
%e.g. \url{https://csgordon.github.io/publications/agere19/}
%I don't want to push to plug my own work if you don't think it makes sense or are making a different point. But I think it's worth
%considering how broadly the big ideas can be applied.
%}

%Advancements in separation logic have been drastic ~\cite{}. These advancements, such as the capability of introducing \emph{ghost} (logical) bookkeeping and associating this bookkeeping with the concept of \emph{invariant} in Iris separation logic, increased the expressivity of any logic that is built on top. Expectedly, many logics enjoying these strong \emph{separation logic} capabilities have been built to verify low-level systems~\cite{}.  %All these program logics built on top of Iris customize and utilize two main aspects of a separation logic like Iris.
%%\paragraph{Custom-Tailored Weakest-Precondition (WP)}
%%\todo[inline]{speak about crash modality etc, encoding how the state should get changed ... GC as well}
%%\paragraph{An Essential Step in Operational Semantics}
%%\todo[inline]{speak about crash step in operational semantics. GC as well}
%%Regarding generalization of customized weakest-preconditions, there has been foundational generalizations defined over incorporating the knowledge of how the resources change, i.e.,transformation of resources, ~\cite{nextgenmodality}, into the separation logic resource assertions. This allows encoding the many customized weakest-preconditions as an instance of a generalized modality definition and lifting of the composition of resource transformation knowledge, which itself is defined as a resource algebra, to the ordinary separation logic resource assertions so that all the underlying resource algebra capabilities can be used for the assertions decorated with the resource-transformation modality. However, even with this kind of generalization effort, it still customizes both the weakest-precondition and resource assertions with the version value (\emph{transformation-generation} value) which most probably lacks being generalizable regarding the real systems to be specified with it as it is not always the case that systems take \emph{versioning} as its key aspect. In general, modalities used in program verification ~\cite{}, are either still tightly coupled with the separation logic they are hosted in, i.e., modalities as definitions for hiding which state a certain truth on resources holds, or as part of the underlying logic's machinery, e.g., later modality as logical machinery encoding step-indexing mechanism.
Modern separation logics are often implemented as frameworks (notably \textsc{Iris}~\cite{jung2018iris}), which offers a collection of tools for building custom separation logic variants for resources of interest, modularly. This makes it possible to
reuse significant amounts of work across verifications using memory~\cite{jung2018iris},
disks~\cite{chajed2019verifying}, and other diverse resources.
%Highly customizable logic ~\cite{iris,..} being the basis for building custom-tailored logics for specific verification tasks ~\cite{some sl refs} enabled building clean and intuitive specifications and proof rules. 
When not using such a framework and instead using an existing tool~\cite{leino2010dafny,vazou2014refinement} as a basis, one can often customize the assertion language by defining libraries of specification definitions and lemmas to effectively have a custom logic.

With these options comes the question: if you can customize your verification tool and how it expresses program states, \emph{how should} you customize it?
We argue that a go-to methodology should be to define \emph{modalities},
which it turns out are in active use in a variety of systems papers because that approach seems to naturally fit many problems that arise in verifying systems code
~\cite{tejperennial19,tejthesis, derekrustbelt20, jung2018iris, kuru2024modalabstractionsvirtualizingmemory, garbagecollect, amalreal2024,fsl, fsl++}.

This paper  
 \begin{itemize}
 \item builds an intuition from a systems-verification perspective about the modal concepts that have been used in recent years with modern program logic -- Section \ref{sec:motivation}
\item describes the common structure underlying the modalities used in systems verification work, effectively a high-level design pattern and process for designing a customized specification language, which has not previously been described --- Section \ref{sec:definitions} % to understand the structure of modal abstractions given for different kinds of systems in Table \ref{table:decomposition}
%\item identifies the common patterns in different modal abstractions for different systems
\item gives example specifications to critical steps of different systems, and discuss how modal abstractions look in these specifications
\end{itemize}

\section{Motivation, Background, and Related Work}
\label{sec:motivation}
To explain our ideas clearly, we briefly recap some of the background and themes our ideas build on,
casting them in a certain way to bring out the relevance of our philosophy.
%\begin{table}[h]
%\centering
%\caption{Decomposing Systems}
%\label{table:decompositionsys}
%\begin{tabular}{@{}lcScP{2cm}@{}}
%\toprule
%   & Column & Column                    & Column                   & Column \\ \midrule
%Row & 1      & 2                         & 3                        & $k=a+b+c+d+e+f+g$      \\
%Row & 5      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8      \\
%Row & 9      & \multicolumn{2}{c}{}                                 & 12     \\
%Row & 13     & $k=a+b+c+d+e$                        & 15                       & 16     \\ \bottomrule
%\end{tabular}
%\end{table}

\subsection{Resources in Systems Software}
\label{sec:systemsoft}
Systems software, in general, interfaces with an underlying computing architecture such that any software system at any higher level in the software stack can (at least indirectly) utilize the resources of the machine.
The last layer of software before the hardware is naturally critical to the correctness of an overall system,
as essentially all software built on top of it assumes its correctness.
%Assuming verified hardware, this makes the correctness of the system software critical to the rest of the software stack. 
%Because they interface with the hardware, abstractions and their implementations are intricate and error-prone. 
And because hardware is complex and highly diverse, the implementation of those lowest layers
of the software stack is typically intricate and naturally error-prone, despite how critical its correctness is.
Typically systems software has, as a primary focus, the task of \emph{abstracting} from hardware details to simplify 
the construction of higher layers of the stack.

\paragraph{Virtualization}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{denning1970virtual}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,rosenblum1992design,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.

\paragraph{Sharing} is one of the fundamental reasons for kernels to {virtualize} resources,
for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.

\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,rosenblum1992design,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees~\cite{}, page tables~\cite{}, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from a system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.

%\paragraph{Isolation Hand-in-Hand with Sharing}
%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%\Colin{Not sure what you want here?}

\subsection{Logic of Resources \& Systems Verification }
\label{sec:seplogic}
Separation logic~\cite{reynolds2002separation} is a specialization of Hoare Logic~\cite{hoare1969axiomatic}
to better deal with reasoning about disjoint resources.
Classic Hoare Logic, as used in prominent systems verifications like \textsc{seL4}~\cite{klein2014comprehensive},
deals with pre- and post-conditions. A so-called ``triple''
\[
\{P\}\;C\;\{Q\}
\]
represents an assertion that if the current program state is accurately described by assertion $P$,
then executing the ``command'' (program) $C$ will not crash, and if it terminates then $Q$ will accurately
describe the ending program state.

\paragraph{Ownership} Separation Logic extends Hoare Logic 
with an additional logical connective $\ast$ (called \emph{separating conjunction}) such that $P \ast Q$ asserts that $P$ and $Q$ not only both hold, but apply to
\emph{disjoint} portions of the program state.\footnote{Though much like systems design, modern separation logics~\cite{jung2018iris}
allow this separation to apply to a kind of virtualization of physical resources~\cite{jensen2012fictional},
or even resources that only exist in the logic~\cite{jung2016higher}.
}
This disjointness assumption leads to discussing separation logic assertions as asserting \emph{ownership}
of the (physical or virtual) resources they hold on.
The classic \emph{points-to} assertion $\ell\mapsto v$ asserts ownership of a portion of memory containing memory address
$\ell$, and that the memory at that location contains value $v$. The disjointness from separating conjunction
is useful: if $\ell\mapsto v \ast \ell'\mapsto v'$ is true, it means $\ell$ and $\ell'$ are distinct memory addresses,
because they must hold on disjoint parts of memory.

Modern separation logics~\cite{jung2018iris} allow definition of more nuanced versions of this idea.
A common form of this, which we revisit in Section \ref{sec:definitions}, is the parameterization of this and
other assertions by a kind of context --- a \emph{resource context} --- resulting in assertions like  $\ell \mapsto_{\textsf{Ctx}} v$, conveying the ownership of the resource defined by the reference $\ell$ and the value ($v$) it's referring to, but where the \emph{resource context}
of the assertion may have the name ($\ell$) and value ($v$) exist not in a partial map of physical memory locations to values,
but of register names to register values~\cite{chlipala2011mostly,ni2007using},
filesystem block indices to block contents~\cite{chajed2019verifying},
or other kinds of associations.
Thus much of the machinery of modern separation logics comes from shifting the notion of which resource context 
assertions are relative to.

%\Colin{Regarding next paragraph: separation logics are not necessarily foundational (early work was all pen-and-paper,
%automated work with SMT solvers clearly isn't foundational). Unless you mean ``foundational'' in a non-technical sense (in which case, a different word should be used, to avoid confusion).}
%\paragraph{Specification} Separation logics~\cite{} rely on foundational Hoare-Logic when introducing their abstractions: make specific assertions and abstractions appear in the form of pre- and post-conditions, $\{P\} e \{Q\}$, called Hoare Triple. It is also important to note that program logics also enjoy encoding Hoare Triples in terms of weakest-precondition~\cite{dijkstra1975guarded} program logics~\cite{pratt1976semantical}, which is roughly
%\[\textsf{P} \; -\ast \;\textsf{wp}\; e \; \{Q\}\]
%where $-\ast$ can  be intuitively thought of as \textit{"when P is added into the resource context we can have the resource represented with Q"}
%\Colin{\emph{Is} it important to note this? I'm genuinely not sure it is.}

\subsection{Modal Reasoning}
\label{sec:modalbackground}
Modal logics are logics of contingent truth, enabling reasoning about cases where something is or may
be true in alternate circumstances. These alternate circumstances are expressed via modal operators,
which decorate a proposition with those alternate circumstances of its truth. This can be express
conditions related to time~\cite{pnueli1977temporal} (e.g., $\diamond P$ for $P$ being true at some point in the future in \textsc{LTL}), necessity and/or possibility~\cite{harel1979first,pratt1976semantical} (e.g., $[C]P$ indicating that $P$ is true after execution
of the program command $C$ terminates), or knowledge and belief~\cite{halpern1990knowledge,hintikka1962knowledge}
(e.g., $\mathcal{K}_a P$ indicating that node $a$ in a distributed system possesses knowledge of $P$),
or the location \emph{where} something is true~\cite{murphy2007type,gordon2019modal}.

\paragraph{Nominals} 
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
This naming generally resembles
the \emph{satisfaction} operator of 
\emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. For this reason we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though the examples we discuss are not necessarily actually hybrid logics.

\subsection{Modal Operators in Program Specifications}
The value of modal operators in program specifications is that they permit describing something that is true
within some context(s), without needing to fully describe the context itself. Yet unlike shorthands,
abbreviations, or general definitions, modal operators interact with the logic in systematic --- rather than ad hoc --- ways.

@@

Although we focus on understanding the modal patterns in the program logics themselves for verifying the client programs (Section \ref{sec:definitions}), not in the logical machinery they implement, we think that it sheds light on how the modal operator utilized in the machinery of the program logics shows resemblance and can be lifted to the task of client verification.

\paragraph{Temporal Operators}
The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
distributed algorithms (e.g., \cite{ongaro2014search,ailijiang2019wpaxos}).

\paragraph{Nominalization}
An example utilization of state naming explicitly on the assertion appears in program logics such as Iris~\cite{jung2018iris}, which enables encoding of usage protocols (e.g. state transition system) resembling typestate~\cite{strom1986typestate,garcia2014foundations} as specifications. Protocol assertions are \emph{annotated with the name of the last (abstract) state at which the protocol is ensured}.
An example more familiar to the systems community is Halpern et al.'s history of adapting modal logics of
knowledge to deal with distributed systems~\cite{halpern1990knowledge,halpern1989modelling,halpern2017reasoning,halpern1990knowledge}.
In most of that line of work, $\mathcal{K}_a(P)$ indicates that the node $a$ in the system \emph{knows} or \emph{possesses knowledge of} $P$
(for example, a Raft node may ``know'' a lower bound on the commit index).
Alternatively, a modality $@_i(P)$ may represent that $P$ is true of/at the specific node $i$~\cite{gordon2019modal} (e.g., that node $i$ has stored a certain
piece of data to reliable storage).
These permit capturing specific concepts relevant to the correctness (and reasoning about correctness) of a certain class of systems,
involving facts about specific named entities in the system.
\\

In each of these cases, the fact that these facts are described using modalities means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion $\mathsf{minCommitIndex}(5)$ to represent the former knowledge,
one would need to separately provide custom reasoning to conclude $\mathsf{minCommitIndex}(3)$.

%\paragraph{Implementing the Logic of Impredicativity} Modal operators such as \emph{later} modality speculating over the fact that modal assertion holds at some state later in the execution are used in the logical machinery of program logics to handle the impredicative terms in the logic. Interestingly, similar reasoning principles using \emph{later} modality are also used in the verification of programs against store-buffers ~\cite{} which also requires speculation over the state in which stored writes appear in the memory.
%
%\paragraph{Persistent Weakest-Precondition} We already gave a rough encoding of Hoare Triple in terms of weakest-precondition, but a thorough version of our definition for the program logics of resources would be 
%\[
%\Box (\textsf{P} \; -\ast \;\textsf{wp}\; e \; \{Q\})
%\]
%decorated with a necessity modality ensures that the weakest-precondition assertion 
%$
%    \textsf{P} \; -\ast \;\textsf{wp}\; e \; \{Q\}
%$
%inside the necessity modality is always valid by the exclusive ownership of the weakest-precondition; as a consequence, the whole specification 
%$
%\Box (\textsf{P} \; -\ast \;\textsf{wp}\; e \; \{Q\})
%$ can be used multiple times -- i.e., a \emph{duplicable resource}.

\section{Logical Decomposition of a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[h]
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
 %$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}cSSSp{5cm}@{}}
\toprule
& Resource Context & Resource Elements  &  Nominalization & Resource Context Steps \\ \midrule
Post-Crash Modality ~\cite{tejthesis, perennialgit,tejperennial19} & $\Diamond \; P $  & $  \ell \mapsto_{n}^{\overline{\gamma}} v $ & Strong &  Crash Recovery   \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
NextGen Modality~\cite{larsnextgen25} &  $\overset{t}{\hookrightarrow} \; P$  & Own (t(a))  & Strong & Determined Based on the Model$^{*}$   \\
StackRegion  Modality$^{*}$  ~\cite{larsnextgen25} & $\overset{ICut^{n}}{\hookrightarrow} \; P$      & $\fbox{n} \; \ell \mapsto v$                                 & Strong & Alloc and Return to/from stack\\
Memory-Fence  Modality ~\cite{fsl,fsl++,derekrustbelt20} & $\triangle_{\pi}$ and $\triangledown_{\pi}$      & $\ell \mapsto v$                                 & Weak & Fence Acquire and Release  \\
Address-Space Modality ~\cite{kuru2024modalabstractionsvirtualizingmemory} & [r]P     &  $\ell \mapsto v$ & Weak                         & Address-Space Switch  \\ 
Ref-Count Modality ~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ & Weak  & Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}

*The StackRegion Modality is an instance of NextGen (called the Independence Modality in \cite{larsnextgen25}).
%and the $\Diamond$ Figure for Post-Crash Modality uses notationis taken from NextGen Modality paper ~\cite{larsnextgen25}.
\end{table*}

In this section, we explain how we can systematically propose new modal abstractions for different types of systems.
We summarize our discussion of these logics in Table \ref{table:decomposition}.
We discuss, based on examples, common aspects of how we intuitively think about correctness of systems code in many contexts,
articulate those pieces, and call out the commonalities across a range of systems.

>>>>

%\bibliography{references/paper}
\begin{document}
%
\title{Modal Verification Patterns for Systems}
%
%\titlerunning{RCU Types}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Ismail Kuru \orcidID{0000-0002-5796-2150} \and
Colin S. Gordon \orcidID{0000-0002-9012-4490}}
%%%Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{I. Kuru et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Drexel University \\ %\and
%%%Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{\{ik335,csgordon\}@drexel.edu}}%\\
%%%\url{http://www.springer.com/gp/computer-science/lncs} \and
%%%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%%%\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
%\maketitle
\section{Introduction}\label{sec:intro}
Low-level systems exhibit certain patterns in their designs, especially when interacting with computing resources. Exploiting certain patterns while designing software has been an important field of study. In this regard, we think that certain properties of modalities enable us to understand and do the verification challenges that show certain patterns.
\subsection{Contributions to Identifying the Verification Patterns via Modal Abstractions}
We argue how modal abstractions can be used to identify and abstract system verification challenges. We justify our perspective by discussing prior systems that have successfully used modalities for system verification, arguing that they fit into the verification design pattern we articulate, and explaining how this approach might apply to other systems' verification challenges.
\paragraph{Identifying System Verification Challenges} We start with identifying common patterns in system verification: \emph{virtualization, sharing, and translation}.
\paragraph{Introducing the Concept of \emph{Resource}} Then we discuss the concept of \emph{resource} which has already been an essential concept in the design of systems. Inspired by the concept of resource in the systems, we define what a resource and its context are in the modal abstractions.
\paragraph{Introducing the Concept of \emph{Nominals}} Nominalization enables identifying a resource in a context. For example, a transaction is a context of resources of in-memory updated disk blocks. The transaction identifier is used to associate a transaction with a disk-block to be persisted so that, in case of a crash while persisting updated disk-blocks, the filesystem can rollback the \emph{already persisted} disk-blocks of the transaction, and reach to the previous consistent disk state. To be able to do so, both the updated in-memory disk blocks and the transaction must refer to the transaction identifier -- \emph{strong nominalization}. In another example, virtual memory references (resources) in an address space (resource context), which can be uniquely identified with a root address (the nominal) of its page-table tree, are \emph{agnostic} to the address space that they are in. However, they can only be accessed (be valid) in the address space to which they are agnostic. However, then an address space switch happens, the virtual memory references of the previous address space must be made inaccessible. To be able to do so, although virtual memory references do not hold any piece of information related to their address space, we still have to associate them. We call this kind of unilateral nominalization -- \emph{weak nominalization}.
\paragraph{Taxonomy of the Current Modal Approaches in System Verification} Based on these concepts defined, we summarize contemporary verification efforts using modal abstractions. We choose them from different domains, for example, weak memory verification, storage persistence ~\cite{tejthesis, perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2024modalabstractionsvirtualizingmemory,amalreal2024} because we would like to justify that our \emph{definitions} are not domain dependent.

\section{Definitions}
Although they differ in the functionality they offer, low-level systems exhibit certain patterns of design and utilization of computing resources.
  In this section, we argue the position that \emph{modalities}  
  should be a go-to approach when specifying and verifying low-level systems code.
  We explain how the concept of a \emph{resource context} helps guide the design of new modalities
  for verification of systems code, and
  we justify our perspective by discussing prior systems that have used modalities for systems verification
  successfully, arguing that they fit into the verification design pattern we articulate,
  and explaining how this approach might apply to other systems verification challenges.

  To explain our ideas in the general systems understanding, we briefly recap some of the background and themes our ideas build on again,casting them in a certain way to bring out the relevance of our philosophy.
  \subsection{Resources in Systems Software}
  \label{sec:systemsoft}
Systems software, in general, interfaces with an underlying computing architecture such that any software system at any higher level in the software stack can (at least indirectly) utilize the resources of the machine.
The last layer of software before the hardware is naturally critical to the correctness of an overall system,
as essentially all software built on top of it assumes its correctness.
%Assuming verified hardware, this makes the correctness of the system software critical to the rest of the software stack. 
%Because they interface with the hardware, abstractions and their implementations are intricate and error-prone. 
And because hardware is complex and highly diverse, the implementation of those lowest layers
of the software stack is typically intricate and naturally error-prone, despite how critical its correctness is.
Typically systems software has, as a primary focus, the task of \emph{abstracting} from hardware details to simplify 
the construction of higher layers of the stack.

\subsubsection{Virtualization}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees~\cite{}, page tables~\cite{}, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from a system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.

%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}

\subsection{Nominals}
\label{sec:modalbackground}

\subsubsection{Recapping Modal Operators in Program Specifications: Systems Perspective}
The value of modal operators in program specifications is that they permit describing something that is true
within some context(s), without needing to fully describe the context itself. Yet unlike shorthands,
abbreviations, or general definitions, modal operators interact with the logic in systematic --- rather than ad hoc --- ways.

@@

Although we focus on understanding the modal patterns in the program logics themselves for verifying the client programs (Section \ref{sec:definitions}), not in the logical machinery they implement, we think that it sheds light on how the modal operator utilized in the machinery of the program logics shows resemblance and can be lifted to the task of client verification. 
\paragraph{Temporal Operators}
The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.
\subsubsection{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
This naming generally resembles
the \emph{satisfaction} operator of 
\emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. For this reason we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though the examples we discuss are not necessarily actually hybrid logics.

An example utilization of state naming explicitly on the assertion appears in program logics such as Iris~\cite{jung2018iris}, which enables encoding of usage protocols (e.g. state transition system) resembling typestate~\cite{strom1986typestate,garcia2014foundations} as specifications. Protocol assertions are \emph{annotated with the name of the last (abstract) state at which the protocol is ensured}.
An example more familiar to the systems community is Halpern et al.'s history of adapting modal logics of
knowledge to deal with distributed systems~\cite{halpern1990knowledge,halpern1989modelling,halpern2017reasoning}.
In most of that line of work, $\mathcal{K}_a(P)$ indicates that the node $a$ in the system \emph{knows} or \emph{possesses knowledge of} $P$
(for example, a Raft node may ``know'' a lower bound on the commit index).
Alternatively, a modality $@_i(P)$ may represent that $P$ is true of/at the specific node $i$~\cite{gordon2019modal} (e.g., that node $i$ has stored a certain
piece of data to reliable storage).
These permit capturing specific concepts relevant to the correctness (and reasoning about correctness) of a certain class of systems,
involving facts about specific named entities in the system.
\\

In each of these cases, the fact that these facts are described using modalities with the core modal property $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$ means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion $\mathsf{minCommitIndex}(5)$ to represent the former knowledge,
one would need to separately provide custom reasoning to conclude $\mathsf{minCommitIndex}(3)$.
\section{Contingency Decomposition of a System}

\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[h]
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}clllp{3.25cm}@{}}
  \toprule
& {\scriptsize Resource Context} & {\scriptsize Resource Elements}  &  {\scriptsize Nominalization} & {\scriptsize Context Steps} \\ \midrule
\scriptsize Post-Crash Modality ~\cite{tejthesis, perennialgit,tejperennial19} & \scriptsize $\Diamond \; P $  & \scriptsize $  \ell \mapsto_{n}^{\overline{\gamma}} v $ & \scriptsize Strong &  \scriptsize Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
\scriptsize NextGen Modality~\cite{larsnextgen25} &  \scriptsize $\overset{t}{\hookrightarrow} \; P$  &\scriptsize  Own (t(a))  & \scriptsize Strong & \scriptsize Determined Based on the Model$^{*}$   \\
\scriptsize StackRegion  Modality$^{*}$  ~\cite{larsnextgen25} & \scriptsize $\overset{ICut^{n}}{\hookrightarrow} \; P$      & \scriptsize $\fbox{n} \; \ell \mapsto v$                                 & \scriptsize Strong & \scriptsize Alloc and Return to/from stack\\
\scriptsize Memory-Fence  Modality ~\cite{fsl,fsl++,derekrustbelt20} & \scriptsize $\triangle_{\pi}$ and $\triangledown_{\pi}$      & \scriptsize $\ell \mapsto v$                                 & \scriptsize Weak & \scriptsize Fence Acquire and Release  \\
\scriptsize Address-Space Modality ~\cite{kuru2024modalabstractionsvirtualizingmemory} & \scriptsize [r]P     &  \scriptsize $\ell \mapsto v$ & \scriptsize Weak                         & \scriptsize Address-Space Switch  \\ 
\scriptsize Ref-Count Modality ~\cite{amalreal2024}&\scriptsize @$_{\ell}$ P    & \scriptsize $\ell_1 \mapsto v$ & \scriptsize Weak  & \scriptsize Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}

{\scriptsize *The StackRegion Modality is an instance of NextGen (called the Independence Modality in \cite{larsnextgen25})}.
%and the $\Diamond$ Figure for Post-Crash Modality uses notationis taken from NextGen Modality paper ~\cite{larsnextgen25}.
\end{table*}

Lots of existing program logics for system verifications have a common structure, which maps to modalities with a couple extra dimensions of design.
We summarize our discussion of these logics in Table \ref{table:decomposition}.
We discuss, based on examples, common aspects of how we intuitively think about correctness of systems code in many contexts,
articulate those pieces, and call out the commonalities across a range of systems.
\subsection{Resource}
Consider first the address-space abstraction in an OS kernel. An address-space of a process is a \textsf{container} of \textsf{virtual} addresses referencing data in memory. One would expect to have \textit{points-to} assertion from separation logic to specify \textit{ownership} of a memory reference pointing to some data. But that ownership is relative to a specific
address space --- a specific container. We tend to think directly about what is true \emph{in an address space},
with the simplest piece being an association between a virtual address and the data it points to.
We call the simplest piece, in this and other examples, the \emph{resource element}:
